Performance analysis of high-resolution \\ ice sheet simulations

### 2022-11-06 paper/perfmod.tex
in form for sending preprints to people
remove from bibliography the references I did not use; move unpublished Ua manual to text

%\documentclass[twocolumn,letterpaper]{igs}
%\documentclass[twocolumn]{igs}  % A4 paper needs no option
\documentclass[review,letterpaper]{igs}

@@

\usepackage[kw]{pseudo}
\pseudoset{left-margin=0mm,topsep=5mm,idfont=\texttt}

@@

In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which fix the velocity following the computation which updates the geometry \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

@@

Certain recent numerical models are semi-implicit in innovative ways.  The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.  However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity is held fixed as the ice thickness is advected.  The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.  (Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain the power $\beta$.)
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015
The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model \citep{Gudmundsson2013,Gudmundsson2021manual} implicitly solves mass continuity free-boundary problem for Equation \eqref{eq:masscontinuity} using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.
% equations (1), (26) in Gudmundson2013
% see semi-implicit comments in commit 9081fdbe3cf250d9cfd66504bf03134043ba2547
% https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m: "Third order impicit Taylor-Galerkin (implict with respect to h, not u and v)"
% https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m: "semi-implicit forward integration. implicit with respect to thickness (h) explicit with respect to velocity"
% and "error('Semi-implicit time integration only implemented for SSTREAM and not SSHEET or Hybrid formulations \n')"

>>>>

%\documentclass[twocolumn]{igs}
\documentclass[review]{igs}
%\documentclass[review,letterpaper]{igs}

@@

@@

In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which fix the velocity following the computation which updates the geometry \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2022}.

@@

Certain recent numerical models are semi-implicit in innovative ways.  The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.  However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity is held fixed as the ice thickness is advected.  The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.  (Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain the power $\beta$.)
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015
The semi-implicit scheme in the open-source \'Ua numerical marine ice sheet model \citep[see also G.~H.~Gudmundsson, \emph{{\'U}a {C}ompendium}, \href{https://github.com/GHilmarG/UaSource}{\texttt{github.com/GHilmarG/UaSource}}]{Gudmundsson2013} implicitly solves mass continuity free-boundary problem for Equation \eqref{eq:masscontinuity} using an active set method, but also while holding the (membrane-stress-resolved) velocity fixed.
% equations (1), (26) in Gudmundson2013
% see semi-implicit comments in commit 9081fdbe3cf250d9cfd66504bf03134043ba2547
% https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m: "Third order impicit Taylor-Galerkin (implict with respect to h, not u and v)"
% https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m: "semi-implicit forward integration. implicit with respect to thickness (h) explicit with respect to velocity"
% and "error('Semi-implicit time integration only implemented for SSTREAM and not SSHEET or Hybrid formulations \n')"
